The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-01-24 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions and additional references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2408 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a4e4e6e9d7ec28bf7326cf7326bc6e6657caf7b5

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with high/extreme variations 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on (a set of) covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is optimized, they differ in their 
specific approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory. All these approaches
have shown to work well for various situations, however, the specific advantages and
disadvantages of each of the testing strategies have not yet been investigated and compared
in detail.

@@

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated.
Moreover, the impact of applying either a pre- or postpruning strategy is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
testing strategies in model-based unbiased recursive partitioning is presented in 
Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then in a separate step the split point. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
\item Using this matrix, the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $Z_j$ on $D$ and in that way the dependency of $Z_j$ on the response $Y$ as the discrepancy
measure (and therefore also $D$) depends on $Y$. 
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
%As the variables are usually not all measured at the same scale the resulting
%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
%variable selection on this standardized/unified scale. 
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

@@

%To emphasize the differences between these three algorithms in how they select split variables,
%their testing strategies are explained in detail in Section~\ref{sec:strategies} 
%using the example of tree models with a linear model fitted in each node as discussed
%in the simulation study. Again, this is done only for means of simplification and the same 
%methodology can be applied for a wide variety of models.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable $Z$.
For different types of response variables different functions $h$ which depend on $Y$ in a 
permutation-symmetric way and might also be multidimensional can be selected.
For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\hat{\beta}$ is the estimated model parameter and $s$ is the score function
$$
s(Y, X, \hat{\beta}) = \frac{\partial \psi(Y, X, \hat{\beta})}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to estimate the model parameter $\beta$. For this purpose many different M-type estimators can 
be applied including maximum likelihood or ordinary least squares. Therefore, these objective 
functions are also possible options to obtain a model-based discrepancy measure for the tree 
algorithm.
Applying the score function $s$ on a data set of $n$ observations containing the 
response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
on a $k$-dimensional estimated parameter vector 
$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix $S$.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
statistic given all permutations of the response variable can be calculated and used to 
standardize the test statistic. In order to map multivariate test statistics into the real 
line two options are given: either the maximum of the absolute values of the standardized 
statistic or a quadratic form is considered. The resulting values can then be transformed to 
the scale of $p$-values where they can be compared in an unbiased way to select a split variable.

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
provides a very general framework where the choice of the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test 
(see \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox).

@@

The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
application of the central limit theorem. 
The fitted model can be of a very basic and simple form such as 
an average value of the response vector
%a constant fit where the parameter $\beta$ is an
%average value of the response, hence 
%$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
%$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ 
but can also be a more complex parametric model.
In \cite{Zeileis+Hothorn+Hornik:2008} GLMs and survival regression models are used to 
exemplify the algorithm. Subsequently various other types of models have been applied as well,
for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.

@@

For a categorical variable all possible splits are evaluated to find the best one while 
for $Z$ being a numeric variable a maximally selected split point search is employed. Hence, 
the number of possible breakpoints is not fixed unlike in the CTree algorithm.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit where the parameter $\beta$ is the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residuals is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as 
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, %(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))
and quantile regression models \citep{Chaudhuri+Loh:2002} among many others.

@@

%\bigskip
%\textit{Note:}
%While GUIDE originally constructs large trees and applies postpruning with the cross-validation 
%method of CART, CTree and MOB employ prepruning by setting a significance level for the 
%independence tests such that no split is performed if the lowest $p$-value is not smaller than 
%the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
%in Section~\ref{sec:simulation} all three algorithms are applied using prepruning in order to put 
%the focus on the power of the algorithms in terms of selecting the best split variable. 
%However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.

@@

\begin{table}[t]
\centering
\caption[Table caption text]{Combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.}
\label{tab:combinations}
\begin{tabular}{ l c c c }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & non-parametric  & conditional inference    & prepruning\\
MOB   & parametric      & score-based fluctuation  & prepruning (or AIC/BIC-based pruning)\\
GUIDE & parametric      & residual-based $\chi^2$  & postpruning (cost-complexity pruning)\\
\hline
\end{tabular}
\end{table}

@@

pn <- partynode(1L, split = sp_xi1, 
                kids = list(
                  #partynode(2L, info = c("true parameters:", 
                  #                       " ",
                  #                       expression(beta[0] == 0~" , "~beta[1] == -delta),
                  #                       "or",
                  #                       expression(beta[0] == +delta~" , "~beta[1] == 1),
                  #                       "or",
                  #                       expression(beta[0] == +delta~" , "~beta[1] == -delta))),
                  #partynode(3L, info = c("true parameters:", 
                  #                       " ",
                  #                       expression(beta[0] == 0~" , "~beta[1] == +delta),
                  #                       "or",
                  #                       expression(beta[0] == -delta~" , "~beta[1] == 1),
                  #                       "or",
                  #                      expression(beta[0] == -delta~" , "~beta[1] == +delta)))))

                  partynode(2L, info = c("true parameters:", 
                                         expression(beta[0] ==~" "~0~" or "~-delta), 
                                         expression(beta[1] ==~" "~1~" or "~+delta))),
                  partynode(3L, info = c("true parameters:", 
                                         expression(beta[0] ==~" "~0~" or "~+delta), 
                                         expression(beta[1] ==~" "~1~" or "~-delta)))))

@@

#par(mfrow = c(2,2))
#par(mar=c(1,1,1,1), oma=c(1,1,1,1))
paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))

@@

%As mentioned before, the GUIDE algorithm usually builds large trees which are then pruned 
%afterwards. However, for the purpose of comparing the testing strategies rather than the whole
%tree algorithms in this simulation study a significance level is considered for all statistical 
%tests, including the $\chi^2$-test in GUIDE such that all algorithms apply prepruning rather 
%than postpruning.
%The effect of these two pruning strategies are investigated in Section~\ref{sec:discussion}.

@@

<<3way_data, results=hide, echo=FALSE>>=
if(file.exists("3way_xi00.rda")){
  load("3way_xi00.rda")
} else {
  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
                delta = 0.3,
                xi = 0, 
                binary_regressor = FALSE,
                binary_beta = TRUE,
                only_intercept = FALSE,
                nrlevels = 2, nrsplitvar = 1,
                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
                return_matrices = TRUE,
                test = c("guide_sum_1_cor",
                         "mfluc_resid_bin",
                         "ctree_resid_bin",
                         "mfluc_resid_cat",
                         "mfluc_resid",
                         "ctree_resid",
                         "guide_sum_12",
                         "mfluc_bin",
                         "ctree_bin",
                         "mfluc_cat",
                         "mfluc",
                         "ctree"))
  d00 <- prep_3way(simres)
  save(d00, file = "3way_xi00.rda")
}

if(file.exists("3way_xi08.rda")){
  load("3way_xi08.rda")
} else {
  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
                delta = 0.5,
                xi = 0.8, 
                binary_regressor = FALSE,
                binary_beta = TRUE,
                only_intercept = FALSE,
                nrlevels = 2, nrsplitvar = 1,
                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
                return_matrices = TRUE,
                test = c("guide_sum_1_cor",
                         "mfluc_resid_bin",
                         "ctree_resid_bin",
                         "mfluc_resid_cat",
                         "mfluc_resid",
                         "ctree_resid",
                         "guide_sum_12",
                         "mfluc_bin",
                         "ctree_bin",
                         "mfluc_cat",
                         "mfluc",
                         "ctree"))
  d08 <- prep_3way(simres)
  save(d08, file = "3way_xi08_delta05.rda")
}
@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.55\textwidth}
<<3way_xi00, fig=TRUE, echo=FALSE, height=4, width=4>>=
par(mar = c(3,5,1,5.7))
with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
                           ylab = expression("Mean"~p~"-value for"~Z[1])))
@
\endminipage
\minipage{0.45\textwidth}
<<3way_xi08, fig=TRUE, echo=FALSE, height=4, width=3.3>>=
par(mar = c(3,2,1,5.8))
with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
                           ylab = expression("Mean"~p~"-value for"~Z[1])))
@
\endminipage
\caption{\label{fig:3way00_08} Mean $p$-values corresponding to the true split variable $Z_1$
for the true split point $\xi=0$ (left panel) and $\xi=0.8$ (right panel),
with both coefficients varying, effect size $\delta = 0.3$, 250 observations and 100 data sets.
All 12 combinations of the three building blocks are evaluated.}
\end{figure}

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<3way_xi00solo, fig=TRUE, echo=FALSE, height=5, width=9.2>>=
par(mar = c(3,5,1,5.7))
with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1,
                           ylab = expression("Mean"~p~"-value for"~Z[1]),
                           xpd = TRUE))
@
\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
for the true split point $\xi=0$,
with both coefficients varying, effect size $\delta = 0.3$, 250 observations and 100 data sets.
All 12 combinations of the three building blocks are evaluated.}
\end{figure}

The results presented in Figure~\ref{fig:3way} confirm the already discussed advantage of
including all available scores rather than just residuals as this leads to a lower mean $p$-value 
of $Z_1$ for all testing strategies.
Moreover, it can be observed that all combinations that binarize scores/residuals perform
worse than those who do not employ any binarization. 
Regardless of the transformation of scores/residuals categorizing split variables also leads
to higher $p$-values and therefore has a negative effect on the power of a testing strategy.
Consequentially, without categorization the $p$-values are lower, particularly for a linear
selection of split variables.

@@

\subsection{Conclusions}
From comparing the effects of the building blocks three main conclusions can be drawn: 
(1)~In most cases including all available score values rather than just residuals leads to 
clearly better results and performs in general at least equally well.
(2)~Binarizing scores/residuals does not lead to any improvement in performance.
(3)~Categorizing possible split variables also doesn't improve the performance 
unless the breakpoints used for the categorization are very close to the true split points.
However, even then the results are very similar to those achieved without any categorization.

%FIX ME: Check implementation of pruning.

%FIX ME: LM-specific results

@@

\section{Applied Combinations of Building Blocks}\label{app:combinations}
For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix represents a change of the original form of the considered testing strategy.
In particular, ``\_res'' refers to using only residuals (instead of all scores),
``\_scores'' to using all scores (instead of only residuals),
``\_cat'' to applying a categorization of split variables,
``\_bin'' to applying a binarization of scores/residuals.

>>>>

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with a strong variation 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is optimized, they differ in their 
specific approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory. All these approaches
have shown to work well for various situations, however, the specific advantages and
disadvantages of each of the testing strategies have not yet been investigated and compared
in detail.

@@

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated.
Additionaly, the impact of applying either a pre- or postpruning strategy is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
testing strategies in model-based unbiased recursive partitioning algorithms is presented in 
Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
\item Using this matrix, the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $Z_j$ on $D$.
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
%As the variables are usually not all measured at the same scale the resulting
%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
%variable selection on this standardized/unified scale. 
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

@@

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice, while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\hat{\beta}$ is the estimated model parameter and $s$ is the score function
$$
s(Y, X, \hat{\beta}) = \frac{\partial \psi(Y, X, \hat{\beta})}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to estimate the model parameter $\beta$. For this purpose many different M-type estimators can 
be applied including maximum likelihood or ordinary least squares. Therefore, these objective 
functions are also possible options to obtain a model-based discrepancy measure for the tree 
algorithm.

Applying the score function $s$ on a data set of $n$ observations containing the 
response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
on a $k$-dimensional estimated parameter vector 
$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix $S$.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
statistic given all permutations of the response variable can be calculated and used to 
standardize the test statistic. In order to map multivariate test statistics into the real 
line two options are given: either the maximum of the absolute values of the standardized 
statistic or a quadratic form. The resulting values can then be transformed to 
the scale of $p$-values where they can be compared in an unbiased way to select a split variable.

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
provides a very general framework where the choice of the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test.
(See \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox.)

@@

The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
application of the central limit theorem. 
The fitted model can be of a very basic and simple form such as an average value of the response 
vector but can also be a more complex parametric model.
In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
are used to exemplify the algorithm. Subsequently various other types of models have been applied as well,
for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.

@@

For a categorical split variable all possible splits are evaluated to find the best one while 
for numeric split variable a maximally selected split point search is employed. Hence, 
the number of possible breakpoints is not fixed unlike in the CTree algorithm.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit where the parameter $\beta$ is the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residuals is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002} 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))

@@

@@

\begin{table}[t]
\centering
\caption[Table caption text]{Combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.}
\label{tab:combinations}
\begin{tabular}{ l l l l }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & non-parametric  & conditional inference    & pre\\
MOB   & parametric      & score-based fluctuation  & pre (or post with AIC/BIC)\\
GUIDE & parametric      & residual-based $\chi^2$  & post (cost-complexity pruning)\\
\hline
\end{tabular}
\end{table}

@@

pn <- partynode(1L, split = sp_xi1, 
                kids = list(
                  partynode(2L, info = c("true parameters:", 
                                         expression(beta[0] ==~" "~0~" or "~-delta), 
                                         expression(beta[1] ==~" "~1~" or "~+delta))),
                  partynode(3L, info = c("true parameters:", 
                                         expression(beta[0] ==~" "~0~" or "~+delta), 
                                         expression(beta[1] ==~" "~1~" or "~-delta)))))

@@

paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))

@@

@@

<<3way_data, results=hide, echo=FALSE>>=
if(file.exists("3way_xi00_delta03.rda")){
  load("3way_xi00_delta03.rda")
} else {
  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
                delta = 0.3,
                xi = 0, 
                binary_regressor = FALSE,
                binary_beta = TRUE,
                only_intercept = FALSE,
                nrlevels = 2, nrsplitvar = 1,
                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
                return_matrices = TRUE,
                test = c("guide_sum_1_cor",
                         "mfluc_resid_bin",
                         "ctree_resid_bin",
                         "mfluc_resid_cat",
                         "mfluc_resid",
                         "ctree_resid",
                         "guide_sum_12",
                         "mfluc_bin",
                         "ctree_bin",
                         "mfluc_cat",
                         "mfluc",
                         "ctree"))
  d00 <- prep_3way(simres)
  save(d00, file = "3way_xi00_delta03.rda")
}

if(file.exists("3way_xi08_delta1.rda")){
  load("3way_xi08_delta1.rda")
} else {
  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
                delta = 1,
                xi = 0.8, 
                binary_regressor = FALSE,
                binary_beta = TRUE,
                only_intercept = FALSE,
                nrlevels = 2, nrsplitvar = 1,
                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
                return_matrices = TRUE,
                test = c("guide_sum_1_cor",
                         "mfluc_resid_bin",
                         "ctree_resid_bin",
                         "mfluc_resid_cat",
                         "mfluc_resid",
                         "ctree_resid",
                         "guide_sum_12",
                         "mfluc_bin",
                         "ctree_bin",
                         "mfluc_cat",
                         "mfluc",
                         "ctree"))
  d08 <- prep_3way(simres)
  save(d08, file = "3way_xi08_delta1.rda")
}

d00$xi <- 0
d08$xi <- 0.8

d3way <- rbind(d00, d08)
d3way$xi <- factor(d3way$xi, levels = c(0,0.8), labels = c(0,0.8))
@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c("red", "blue", "red", "blue"),
       lty = c(1,1,2,2),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
                                              expression(xi==0.8~"and"~delta==1))),
       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
                  lines = list(lty = c(1,1,2,2), 
                               type = "l",
                               col = c("red", "blue", "red", "blue"))))

#par(mar = c(3,5,1,5.7))
#with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
#                           ylab = expression("Mean"~p~"-value for"~Z[1])))
#with(d00, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d00, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
#with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1)))
#with(d08, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d08, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
@
\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
over different levels of the building block \textit{Categorization}. 
Each line represents
a combination of the levels of the other two building blocks \textit{Binarization} and 
\textit{Residuals vs.\ Scores}.
Left panel: true split point $\xi=0$ and effect size $\delta = 0.3$.
Right panel: true split point $\xi=0.8$ and effect size $\delta = 1$.}
\end{figure}

The results presented in Figure~\ref{fig:3way} clearly show that binarizing scores/residuals
reduces the power of the testing strategy as it leads to higher $p$-values corresponding
to the true split variable $Z_1$. 
This effect can be observed for a central split point (left panel, $xi=0$) as well as for a
a split point closer to upper end of the considered interval (right panel, $xi=0.8$).
However, the effect of the building block \textit{Categorization} varies over these two settings.
For a central split point all three levels perform similarly. The linear selection of split variables
is slightly ahead even though the true parameter function is a step function. 
On the contrary, clear differences can be seen for the outward split point. Here, the maximal selection
performs best while the linear selection shows its disadvantage in detecting abrupt changes
and the categorized split variables lead to even higher $p$-values and thus lower power.
Overall, it depends on the situation whether a linear or a maximal selection of a split variable
is advantageous. However, categorizing possible split variables does not lead to any improvements
but even declines the performance of a testing strategy unless the true split point is close
to a break point used for the binarization.

@@

\fixme
From comparing the effects of the building blocks three main conclusions can be drawn: 
(1)~In most cases including all available score values rather than just residuals leads to 
clearly better results and performs in general at least equally well.
(2)~Binarizing scores/residuals does not lead to any improvement in performance.
(3)~Categorizing possible split variables also doesn't improve the performance 
unless the breakpoints used for the categorization are very close to the true split points.
However, even then the results are very similar to those achieved without any categorization.

@@

\section{Applied Combinations of Building Blocks}\label{app:combinations}
For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``\_res'' refers to using only residuals (instead of all scores),
``\_scores'' to using all scores (instead of only residuals),
``\_cat'' to applying a categorization of split variables,
``\_bin'' to applying a binarization of scores/residuals.
